\section{Abstract syntax tree}

All the parsers implement the design idea (see section XX \fixme{ref to design})
of only having one data type for the possible nodes in the AST. 

The MLB and SML AST is defined as a tree of wrapped nodes. We recall that the
wrap type is a record with a left, right and node field, where the left and
right fields are of the same type:

\begin{sml}
signature Wrap =
sig
  type ('a, 'b) t = {node: 'a, left: 'b, right: 'b}
  ...
end
\end{sml}

The node field stays the same for both AST's but the type of the left and right
fields change over time as the AST is resolved or transformed by tools.

The MLB parser returns unit in the left and right wrap fields and the SML parser
returns the nodes left and right integer position in the source code as the left
and right wrap fields respectively.

Right now the left and right fields of the MLB AST are not used (only set to
unit), but the idea is that the environment, as it looks prior to the SML or MLB
file, should be saved such that if the file changes we only need to re parse the
changed part and can then use the environment as it looked before this
file. This way we don't need to re parse the whole MLB file (most likely
including a re parse of the basis library). This has however not been implement
as of this writing (see \fref{sec:future-work} for discussion about space issues
where this could help out).

\subsection{Resolving}

The two implemented resolvers are working purely on the SML AST and they share
the same idea of the left field being how it all looks before the node and the
right is what the node contributes/changes.

\subsection{Infix resolving}

The infix resolver adds an infix basis to every node and changes all
\ttt{Exp_FlatApp} and \ttt{Pat_FlagApp} to \ttt{Exp_App} and \ttt{Pat_app}
respectively. The right hand infix basis is always empty unless the node is an infix
declaration.

The infix basis is a dictionary of variable names to their fixity status (left,
right or nonfix).

\subsection{Environment resolving}

\fixme{insert brief description of what is storred and the unique numberings}


 

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../report"
%%% End: 
